---
title: "STA322 - Project 1"
author: "Julia Zheng and Christina Zhang"
editor: source
format: pdf
---

```{r}

library(survey)
library(tidyverse)

faculty <- read.csv("trinity faculty.csv")
```

-   stratified sampling with each department as a stratum (41 in total and labeled in alphabetical order 1 through 41)

-   proportional allocation calculated for each and rounded to the nearest whole number

    -   values rounded down to 0 were manually changed to 1 (insert justification)

-   faculty IDs are created based on their order of appearance on the [Duke Trinity Scholars](https://scholars.duke.edu/organization/) website, in surname alphabetical order

```{r}
#| label: renaming-variables

faculty <- faculty |>
  rename(
    stratum = 'Stratum..',
    department = 'Dept.name',
    N_h = 'X..of.faculty',
    n_h = 'Rounded.allocation'
  ) |>
  arrange(stratum)
```

```{r}
#|label: create-facultyIDs

trin_faculty <- faculty |>
  rowwise() |>
  reframe(
    stratum = stratum,
    department = department,
    faculty_id = 1:N_h,
    n_h = n_h
  )
```

```{r}
#|label: stratified-random-sample

set.seed(215)

sample <- trin_faculty |>
  group_by(stratum, department) |>
  group_modify(~ slice_sample(.x, n = .x$n_h[1], replace = FALSE)) |> 
  #for julia: fyi this line of code debugged by chat
  ungroup() |>
  arrange(stratum, faculty_id)

sample
```

Question 1. What is an estimate of the fraction of faculty who are women?

Question 2. What fraction of faculty have at least one grant that is ongoing in 2026?

Question 3. Among faculty who have a Ph.D. degree, what is the average number of years since they received the Ph.D. degree?

Question 4. Question of our choice
