---
title: "STA322 - Project 1"
author: "Julia Zheng and Christina Zhang"
editor: source
format: pdf
---

# Introduction

```{r}
library(survey)
library(tidyverse)
library(pps)
library(dplyr)
```

```{r}
# load and reorganize data

trinity <- read_csv("trinity_faculty.csv") 

trinity <- trinity |>
  select(department = "Dept name", M_i = "# of faculty")

# certainty departments (pi_i = 1), same 7 as in the sampling step
certainty_depts <- c("Economics", "Physics", "Romance Studies", "Music",
                     "Biology", "Mathematics", "Psychology & Neuroscience")

certainty <- trinity |>
  filter(department %in% certainty_depts) |>
  mutate(pi_i = 1)

# remaining departments: 13 more selected, pi_i = 13 * M_i / sum(M_i)
remaining <- trinity |>
  filter(!department %in% certainty_depts)
remaining$pi_i <- 13 * remaining$M_i / sum(remaining$M_i)

max(remaining$pi_i)   # should be 0.9246, matching the sampling step

dept_info <- bind_rows(certainty, remaining)
```

```{r}
# systematic PPS sampling
n_dept <- 20

t_x <- sum(trinity$M_i) # total number of faculty across all departments

trinity$pi_i <- n_dept * trinity$M_i / t_x # first-order inclusion probability for each department

max(trinity$pi_i) # check assumption that nxi/tx <= 1 for all i. it's not...

# check which department pis > 1 --> treat as certainty cases
trinity |>
  filter(pi_i > 1) |>
  select(department, M_i, pi_i)

```

```{r}
# dealing with certainty cases

# separate initial certainty and non-certainty departments
certainty <- trinity |>
  filter(pi_i >= 1) |>
  mutate(pi_i = 1)

remaining <- trinity |>
  filter(pi_i < 1)

n_remaining <- n_dept - nrow(certainty) # calculate number of additional departments to select

remaining$pi_i <- n_remaining * remaining$M_i / sum(remaining$M_i) # recalculate pi_i for remaining departments

# if recalculation creates additional certainty PSUs, repeat until all remaining probabilities are below 1
while (any(remaining$pi_i >= 1)) {
  new_certainty <- remaining |>
    filter(pi_i >= 1) |>
    mutate(pi_i = 1)
  certainty <- bind_rows(certainty, new_certainty)
  remaining <- remaining |>
    filter(pi_i < 1)
  n_remaining <- n_dept - nrow(certainty)
  remaining$pi_i <- n_remaining * remaining$M_i / sum(remaining$M_i)
}

# results check
certainty |>
  select(department, M_i, pi_i) # 7 certainty cases identified (1 additional throughout recalculations)

nrow(certainty)
n_remaining
max(remaining$pi_i)
```

```{r}
set.seed(919)
# select remaining cases with pps & add certainty cases

selected_depts <- ppss(remaining$M_i, n_remaining) # returns numbers of selected rows

pps_remaining <- remaining[selected_depts, ] # keep selected departments

pps_sample <- bind_rows(certainty, pps_remaining) # combine certainty psus with pps-selected departments

nrow(pps_sample) # check 20 departments were selected

# view selected departments
pps_sample |>
  select(department, M_i, pi_i)
```

```{r}
# stage 2 srs

# srs 5 faculty from each department
pps_sample$faculty_numbers <- lapply(
  pps_sample$M_i,
  function(M_i) sample(1:M_i, 5)
)

# view numbers of faculty selected
pps_sample |>
  select(department, M_i, pi_i, faculty_numbers)
```

```{r}

samples <- read.csv("pps_sample.csv") 

samples <- samples |>
  select(-any_of(c("M_i", "pi_i", "pi", "weight"))) |>
  rename(
    woman = Sex..1...F..0...other.,
    grant = Ongoing.grant...1...Yes..0...No.,
    has_phd = PhD...1...Yes..0...No.,
    yrs_phd = Number.of.years.since.PhD,
    yrs_pub = Years.since.last.pub,
    intl_ugrad = International.undergrad.degree.
  )

# 3. attach M_i and pi_i, then build the weight
samples <- samples |>
  left_join(dept_info, by = "department")

samples$weight <- (1 / samples$pi_i) * (samples$M_i / 5)

# 4. survey design
fac_des <- svydesign(
  ids = ~department,
  weights = ~weight,
  data = samples
)
```

Question 1. What is an estimate of the fraction of faculty who are women?

```{r}
#|label: women-prop

prop_women <- svymean(~woman, fac_des)
prop_women

confint(prop_women, level = .95)

```

Question 2. What fraction of faculty have at least one grant that is ongoing in 2026?

```{r}
#|label: grant

prop_grant <- svymean(~grant, fac_des)
prop_grant

confint(prop_grant, level = .95)

```

Question 3. Among faculty who have a Ph.D. degree, what is the average number of years since they received the Ph.D. degree?

```{r}
#|label: phd

phd_des <- subset(fac_des, has_phd == 1)

avg_phd <- svymean(~yrs_phd, phd_des)
avg_phd

confint(avg_phd, level = .95)

```

Question 4.

a\) Among those who have published papers, what is the avg number of years since their last publication?

```{r}
#|label: pub-yrs

pub_des <- subset(fac_des, !is.na(yrs_pub))

avg_pub <- svymean(~yrs_pub, pub_des)
avg_pub

confint(avg_pub, level = .95)

```

b\) Did this faculty receive their undergraduate degree outside of the US?

```{r}
#|label: int-ug

prop_intl <- svymean(~intl_ugrad, fac_des)
prop_intl

confint(prop_intl, level = .95)

```
