---
title: "STA322 - Project 1"
author: "Julia Zheng and Christina Zhang"
editor: source
format: pdf
---

# Introduction

hi

```{r}
library(survey)
library(tidyverse)
library(pps)
```

- PPS sampling without replacement was used, with each of the 41 Trinity departments treated as a primary sampling unit.

- A department’s measure of size, (M_i), was its number of faculty. Thus, departments with more faculty had a larger probability of selection.

- We aimed for a sample size of 100, so we chose to do a two-stage PPS cluster sample from 20 departments and do a SRS of 5 faculty from each department selected.

- A seed of 215 was set so that the sample can be reproduced.

- Although departments were selected without replacement, we calculated the with-replacement variance estimator. This treats the selected departments as if they had been independently sampled with replacement and provides a conservative variance approximation.

```{r}
# load and reorganize data

trinity <- read_csv("trinity_faculty.csv")

trinity <- trinity |>
  select(stratum = "Stratum #", 
         department = "Dept name", 
         M_i = "# of faculty")

n_dept <- 20 # number of departments to select

trinity <- trinity[sample(1:nrow(trinity)), ] # randomize order of rows since systematic sampling depends on order

```

```{r}
# systematic PPS sampling

t_x <- sum(trinity$M_i) # total number of faculty across all departments

trinity$pi_i <- n_dept * trinity$M_i / t_x # first-order inclusion probability for each department

max(trinity$pi_i) # check assumption that nxi/tx <= 1 for all i. it's not...

# check which department pis > 1 --> treat as certainty cases
trinity |>
  filter(pi_i > 1) |>
  select(department, M_i, pi_i)

```

```{r}
# dealing with certainty cases

# separate initial certainty and non-certainty departments
certainty <- trinity |>
  filter(pi_i >= 1) |>
  mutate(pi_i = 1)

remaining <- trinity |>
  filter(pi_i < 1)

n_remaining <- n_dept - nrow(certainty) # calculate number of additional departments to select

remaining$pi_i <- n_remaining * remaining$M_i / sum(remaining$M_i) # recalculate pi_i for remaining departments

# if recalculation creates additional certainty PSUs, repeat until all remaining probabilities are below 1
while (any(remaining$pi_i >= 1)) {
  new_certainty <- remaining |>
    filter(pi_i >= 1) |>
    mutate(pi_i = 1)
  certainty <- bind_rows(certainty, new_certainty)
  remaining <- remaining |>
    filter(pi_i < 1)
  n_remaining <- n_dept - nrow(certainty)
  remaining$pi_i <- n_remaining * remaining$M_i / sum(remaining$M_i)
}

# results check
certainty |>
  select(department, M_i, pi_i) # 7 certainty cases identified (1 additional throughout recalculations)

nrow(certainty)
n_remaining
max(remaining$pi_i)
```

```{r}
set.seed(919)
# select remaining cases with pps & add certainty cases

selected_depts <- ppss(remaining$M_i, n_remaining) # returns numbers of selected rows

pps_remaining <- remaining[selected_depts, ] # keep selected departments

pps_sample <- bind_rows(certainty, pps_remaining) # combine certainty psus with pps-selected departments

nrow(pps_sample) # check 20 departments were selected

# view selected departments
pps_sample |>
  select(department, M_i, pi_i)
```

```{r}
# stage 2 srs

# srs 5 faculty from each department
pps_sample$faculty_numbers <- lapply(
  pps_sample$M_i,
  function(M_i) sample(1:M_i, 5)
)

# view numbers of faculty selected
pps_sample |>
  select(department, M_i, pi_i, faculty_numbers)
```

Question 1. What is an estimate of the fraction of faculty who are women?

Question 2. What fraction of faculty have at least one grant that is ongoing in 2026?

Question 3. Among faculty who have a Ph.D. degree, what is the average number of years since they received the Ph.D. degree?

Question 4. Question of our choice

OLD STRATIFIED SAMPLING CODE

- stratified sampling with each department as a stratum (41 in total and labeled in alphabetical order 1 through 41)

- proportional allocation calculated for each and rounded to the nearest whole number

  - values rounded down to 0 were manually changed to 1 (insert justification)

- faculty IDs are created based on their order of appearance on the [Duke Trinity Scholars](https://scholars.duke.edu/organization/) website, in surname alphabetical order
